|

### 2020-07-28 paper.Rmd
Updated notation.

To understand how the $p$ dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius. This dependence was desribed in @Laa:2020wkm. We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r, R, p)$, where $r$ can be any radius within $[0, R]$. The relative projected volume is then given as the ratio of these two quantities,
\begin{equation}
v_{2} (r, p, R) = \frac{V_{2D}(r, R, p)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}
This ratio is of particular interest because it gives the 2-dimensional radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$-dimensional hypersphere.

We can compare $v_{2} (r, p, R)$ to the relative volume within a radius $r$ in the original $p$-dimensional hypersphere,
\begin{equation}
v_{p} (r, p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
\end{equation}
This comparison is shown in Fig. \ref{fig:cdf}, for $p=3, 10, 100$. The left graph shows $v_{p} (r, p, R)$ and the right graph $v_{2} (r, p, R)$. We clearly see the opposing behavior of the two functions as $p$ increases. While $v_{p} (r, p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r, p, R)$ has the inverse behavior.

@@

The aim of the algorithm is to redistribute the projected volume such that equal relative areas on the disk, as given by $v_{2} (r, p=2, R)= v_p(r, p=2, R) = (r/R)^2$, contain equal relative projected volume, given by $v_{2} (r, p, R)$. This is achieved through a transformation of the projected radius that can be defined for any $r\in[0,R]$, and is applied to the projected data points in the plane, $x = (x_1, x_2)$. We work with polar coordinates and represent the data points as $x = (r_x, \theta_x)$. The angular component $\theta_x$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed. The radial component $r_x$ is transformed in two steps.

The first transformation is to replace $r_x$ with $v_{2} (r_x, p, R)$. Since this is the radial CDF of the assumed underlying distribution, we expect that $v_{2} (r_x, p, R)$ is approximately uniformly distributed in radius. We then transform $v_{2} (r_x, p, R)$ using the inverse of $v_{2} (r_x, 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk. This inverese is defined via
\begin{equation}
v_2(v_2^{inv}(r_x, 2, R), 2, R) = \left(\frac{v_2^{inv}(r_x, 2, R)}{R}\right)^2= r_x
\end{equation}
and thus
\begin{equation}
v_2^{inv}(r_x, 2, R) = R \sqrt{r_x}.
\end{equation}

The full radial transformation is therefore given by 
\begin{equation}
r'_x = v_2^{inv} (v_2(r_x, p, R), 2, R) =  R \sqrt{v_2(r_x, p, R)} = R \sqrt{1-\left(1-\left(\frac{r_x}{R}\right)^2\right)^{p/2}}.
\label{eq:resc}
\end{equation}

The relation between $r'_x$ and $r_x$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}. We see that the transformation is approximately linear near the center. As $p$ increases it becomes non-linear faster, and e.g. for $p=10$ the points with radius $r_x>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'_x$. Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq. \ref{eq:resc}.

```{r radii, fig.cap="Relation between $r_x$ and $r'_x$ for different values of $p$ and assuming $R=1$. The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
# define index as function of c
trans_p <- function(p){
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

# plot dependence for selected values
ggplot(data = data.frame(x = c(0,1)), mapping = aes(x = x)) +
    stat_function(
        fun = trans_p(2),
        mapping = aes(color = "ca")) +
  stat_function(
        fun = trans_p(4),
        mapping = aes(color = "cb")) +
  stat_function(
        fun = trans_p(6),
        mapping = aes(color = "cc")) +
    stat_function(
        fun = trans_p(10),
        mapping = aes(color = "cd")) +
    stat_function(
        fun = trans_p(20),
        mapping = aes(color = "ce")) +
  scale_color_manual(name = "p",
                     values = RColorBrewer::brewer.pal(5, "Dark2"),
                     labels = c("2 (baseline)", "4", "6", "10", "20")) +
  xlab(expression(r[x])) + ylab(expression(r[x]*minute)) + theme_bw()

@@

We therefore allow trimming of the projected observations, using $R$ as a free parameter of the display function. When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$, by trimming $r_x$ as
\begin{equation}
r_x^{\mathrm{trim}} = \min(r_x, R)
\label{eq:cutR}
\end{equation}
for each observation $x$.

@@

We allow tuning $p_{\mathrm{eff}} = s p$ by selecting the scaling parameter $s$. By default, $s=1$ and $p_{\mathrm{eff}}=p$. When $s<1$ the rescaling will be softer, and $s>1$ results in more aggressive rescaling than suggested by $p$ alone. Note that when $p_{\mathrm{eff}} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.

```{r scaled, eval=F, fig.cap="Relation between $r$ and $r'$ for $p=10$ and different values of the scaling $s$. We can tune $s$ to control how strong the rescaling is, with $s<1$ leading to a weaker effect, and large values of $s$ increase the effect.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
# define index as function of c
trans_s <- function(s, p){
  p <- s * p
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

# plot dependence for selected values
ggplot(data = data.frame(x = c(0,1)), mapping = aes(x = x)) +
    stat_function(
        fun = trans_s(0.2, 10),
        mapping = aes(color = "ca")) +
  stat_function(
        fun = trans_s(0.5, 10),
        mapping = aes(color = "cb")) +
  stat_function(
        fun = trans_s(1, 10),
        mapping = aes(color = "cc")) +
    stat_function(
        fun = trans_s(2, 10),
        mapping = aes(color = "cd")) +
    stat_function(
        fun = trans_s(5, 10),
        mapping = aes(color = "ce")) +
  scale_color_manual(name = "s",
                     values = RColorBrewer::brewer.pal(5, "Dark2"),
                     labels = c("0.2", "0.5", "1", "2", "5")) +
  xlab("r") + ylab("r'") + theme_bw()

@@

1. Obtain projection matrix $\mathbf{A}$. For $d$-dimensional projections this is an orthonormal $p \times d$ matrix. To ensure the smooth rotation of projections, each new $\mathbf{A}$ is obtained as an interpolated step in the sequence, as explained in @BCAH05.
2. Project the data by computing $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$.
3. Map $\mathbf{Y}$ to the display to re-draw the projected data. For $d=2$ this typically maps the projected points onto a scatter plot display. With the new display we first transform $\mathbf{Y}$ as:
    + Center the 2-dimensional matrix $\mathbf{Y}$ and compute its polar coordinate representation $(r_y, \theta_y)$.
    + For each observation, first use Eq. \ref{eq:cutR} to get the trimmed radius $r_y^{\mathrm{trim}}$  within the specified range, and then apply the radial transformation defined in Eq. \ref{eq:resc} to obtain $r'_y$.
    + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y_1, y_2)$.
4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using the `half_range` parameter.

@@

```{r codeExample, echo=TRUE, eval=FALSE}
tourr::animate(
  data,
  tour_path = tourr::grand_tour(),
  display = display_sage(s, R, half_range)
  )
```

and introduces the scaling parameter `s` for computing $p_{\mathrm{eff}}$, and the overall scale `R` used for trimming. Both these parameters are described in Section \ref{sec:params} above. The ratio of `half_range` and `R` sets the scale for fitting the displayed data on the plotting canvas, by default `half_range` = `R` and we apply the scaling factor $0.9$ to contain the projected points within the display. When adjusting `R` the user should take care to adjust the `half_range` parameter accordingly.

>>>>

To understand how the $p$ dimensional volume is projected onto a $2$ dimensional plane, we study what fraction of the total volume is projected onto the area of a disk depending on its radius. This dependence was desribed in @Laa:2020wkm. We start from a $p$ dimensional hypersphere with radius $R$ and volume $V(R, p)$, and its projected volume onto a centered $2$-dimensional disk of radius $r$, $V_{2D}(r; p, R)$, where $r$ can be any radius within $[0, R]$. The relative projected volume is then given as the ratio of these two quantities,
\begin{equation}
v_{2} (r; p, R) = \frac{V_{2D}(r; p, R)}{V(R, p)} = 1 - \left(1-\left(\frac{r}{R}\right)^2\right)^{p/2}.
\label{eq:cdf}
\end{equation}
This ratio is of particular interest because it gives the 2-dimensional radial cumulative distribution function (CDF) of points when assuming a uniform distribution within the $p$-dimensional hypersphere.

We can compare $v_{2} (r; p, R)$ to the relative volume within a radius $r$ in the original $p$-dimensional hypersphere,
\begin{equation}
v_{p} (r; p, R) = \frac{V(r, p)}{V(R, p)} = \left({\frac{r}{R}}\right)^p.
\end{equation}
This comparison is shown in Fig. \ref{fig:cdf}, for $p=3, 10, 100$. The left graph shows $v_{p} (r; p, R)$ and the right graph $v_{2} (r; p, R)$. We clearly see the opposing behavior of the two functions as $p$ increases. While $v_{p} (r; p, R)$ concentrates more and more towards large values of $r/R$, $v_{2} (r; p, R)$ has the inverse behavior.

@@

The aim of the algorithm is to redistribute the projected volume such that equal relative areas on the disk, as given by $v_{2} (r; p=2, R)= v_p(r; p=2, R) = (r/R)^2$, contain equal relative projected volume, given by $v_{2} (r; p, R)$. This is achieved through a transformation of the projected radius that can be defined for any $r\in[0,R]$, and is applied to the projected data points in the plane, $y = (y_1, y_2)$. We work with polar coordinates and represent the data points as $y = (r_y, \theta_y)$. The angular component $\theta_y$ is uniform for this distribution, by the rotation invariance of the sphere, and thus does not need to be transformed. The radial component $r_y$ is transformed in two steps.

The first transformation is to replace $r_y$ with $v_{2} (r_y; p, R)$. Since this is the radial CDF of the assumed underlying distribution, we expect that $v_{2} (r_y; p, R)$ is approximately uniformly distributed in radius. We then transform $v_{2} (r_y; p, R)$ using the inverse of $v_{2} (r_y; 2, R)$, to go from a uniform distribution in radius to a uniform distribution in area of the disk. This inverese is defined via
\begin{equation}
v_2^{-1}(v_2(r_y; 2, R); 2, R) = v_2(v_2^{-1}(r_y; 2, R); 2, R) = r_y
\end{equation}
and thus
\begin{equation}
v_2^{inv}(r_y; 2, R) = R \sqrt{r_y}.
\end{equation}

The full radial transformation is therefore given by 
\begin{equation}
r'_y = v_2^{inv} (v_2(r_y; p, R); 2, R) =  R \sqrt{v_2(r_y; p, R)} = R \sqrt{1-\left(1-\left(\frac{r_y}{R}\right)^2\right)^{p/2}}.
\label{eq:resc}
\end{equation}

The relation between $r'_y$ and $r_y$ depends on the number of dimensions $p$, and is illustrated for selected values in Figure \ref{fig:radii}. We see that the transformation is approximately linear near the center. As $p$ increases it becomes non-linear faster, and e.g. for $p=10$ the points with radius $r_x>0.5$ will already be highly distorted and pushed out towards the last eighth in $r'_x$. Figure \ref{fig:circles} demonstrates this for different values of $p$ by showing equidistant circles for which the radius has been transformed according to Eq. \ref{eq:resc}.

```{r radii, fig.cap="Relation between $r_y$ and $r'_y$ for different values of $p$ and assuming $R=1$. The scaling is approximately linear near the center, but leads to distortion at large radii when $p$ is large.", fig.height=3, fig.width=6, out.width="60%", fig.align = "center"}
# define index as function of c
trans_p <- function(p){
  function(r){
    sqrt( 1 - (1-r^2)^(p/2))
  }
}

# plot dependence for selected values
ggplot(data = data.frame(x = c(0,1)), mapping = aes(x = x)) +
    stat_function(
        fun = trans_p(2),
        mapping = aes(color = "ca")) +
  stat_function(
        fun = trans_p(4),
        mapping = aes(color = "cb")) +
  stat_function(
        fun = trans_p(6),
        mapping = aes(color = "cc")) +
    stat_function(
        fun = trans_p(10),
        mapping = aes(color = "cd")) +
    stat_function(
        fun = trans_p(20),
        mapping = aes(color = "ce")) +
  scale_color_manual(name = "p",
                     values = RColorBrewer::brewer.pal(5, "Dark2"),
                     labels = c("2 (baseline)", "4", "6", "10", "20")) +
  xlab(expression(r[y])) + ylab(expression(r[y]*minute)) + theme_bw()

@@

We therefore allow trimming of the projected observations, using $R$ as a free parameter of the display function. When selecting a value $R$ that is smaller than the maximum distance from the center, we need to ensure that the projected radius of points is always smaller than $R$, by trimming $r_y$ as
\begin{equation}
r_y^{\mathrm{trim}} = \min(r_y, R)
\label{eq:cutR}
\end{equation}
for each observation $x$.

@@

We allow tuning $p_{\mathrm{eff}} = \gamma p$ by selecting the scaling parameter $\gamma$. By default, $\gamma=1$ and $p_{\mathrm{eff}}=p$. When $\gamma<1$ the rescaling will be softer, and $\gamma>1$ results in more aggressive rescaling than suggested by $p$ alone. Note that when $p_{\mathrm{eff}} < 2$ we actually invert the behavior and shift the focus away from the center, in general this is not wanted.

@@

1. Obtain projection matrix $\mathbf{A}$. For $d$-dimensional projections this is an orthonormal $p \times d$ matrix. To ensure the smooth rotation of projections, each new $\mathbf{A}$ is obtained as an interpolated step in the sequence, as explained in @BCAH05.
2. Project the data by computing $\mathbf{Y} = \mathbf{X}\cdot\mathbf{A}$.
3. Map $\mathbf{Y}$ to the display to re-draw the projected data. For $d=2$ this typically maps the projected points onto a scatter plot display. With the new display we first transform $\mathbf{Y}$ as:
    + Center the 2-dimensional matrix $\mathbf{Y}$ and compute its polar coordinate representation $(r_y, \theta_y)$.
    + For each observation, first use Eq. \ref{eq:cutR} to get the trimmed radius $r_y^{\mathrm{trim}}$  within the specified range, and then apply the radial transformation defined in Eq. \ref{eq:resc} to obtain $r'_y$.
    + Use the transformed radial coordinate $r'_y$ to re-compute the mapping onto Euclidean coordinates $(y_1, y_2)$.
4. To fit the final projection onto the plotting canvas ranging between $[-1,1]$, we rescale each mapped observation $y$ using a scaling parameter $s$.

@@

```{r codeExample, echo=TRUE, eval=FALSE}
tourr::animate(
  data,
  tour_path = tourr::grand_tour(),
  display = display_sage(gam, R, half_range)
  )
```

and uses `gam` to set the $\gamma$ parameter for computing $p_{\mathrm{eff}}$, and the overall range `R` used for trimming. Both these parameters are described in Section \ref{sec:params} above. Finally, `half_range` sets the scale parameter $s$ to adjust the scale to the drawing canvas. The ratio $R/s$ sets the scale for fitting the displayed data on the plotting canvas, by default $s = R$ and we apply the scaling factor $0.9$ to contain the projected points within the display. When adjusting $R$ the user should take care to adjust $s$ accordingly.
